Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-23 qnml.tex
edited qNML-section

Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
one for each variable, but unlike the other two, it is not based on
further partitioning the corresponding data column
\begin{eqnarray}
s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}
                             {P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
\end{eqnarray}
The trick here is to model a subset of columns as though there were no
conditional independencies among the corresponding variables $S
\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
value configurations and consider them as values of a single variable
with $\prod_{X_i\in S} r_i$ different values which can then be modeled
with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
necessarily define a distribution for $D$, but it is easy to verify
that it coincides with the $P_{NML}(D;G)$ for all the networks that
are composed of fully connected components.  The number of such
networks equals the number of nonempty partitions of a set of $n$
elements, i.e., the $n^\text{th}$ Bell number.

@@

The qNML yields equal scores for network structures that encode same
sets of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those who a) are same when directed
arcs are substitued by undirected ones and b) who have same
V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
$B$ are parents of $C$, but there is no arc between $A$ and $B$ (in
neither direction).  Later, Chickering~\cite{Chick95} showed that all
the equivalent network structures, and only them, can be reached from
each other by reversing, one by one, so called covered arcs, i.e. the
arcs from node $A$ to $B$, for which $B$'s parents other than $A$ are
exactly the $A$'s parents ($G_B=\{A\}\cup G_A$).

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
suplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notations and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  due to equations
  $\{A\}\cup G_A  = G_B$,
  $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
  and $G_A        = G'_B$.
\end{proof}

@@

The BIC can be written as
\begin{equation}
\textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
\end{equation}where $D_i = D[\cdot, i]$ stands for the data on variable $i$, $\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of the conditional distribution of variable $i$ given its parents in $G$. Since the both BIC and qNML scores are decomposable, we can focus on studying the local scores. Using the definition of qNML, we have
\begin{align}\label{qnmlDef2}
s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\
& -(reg(N,q_i r_i) - reg(N,q_i)).
\end{align}

@@

>>>>

Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
one for each variable, but unlike the other two, it is not based on
further partitioning the corresponding data column
\begin{eqnarray}
\label{eqn:qnmldef}
s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}
                             {P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
\end{eqnarray}
The trick here is to model a subset of columns as though there were no
conditional independencies among the corresponding variables $S
\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
value configurations and consider them as values of a single variable
with $\prod_{X_i\in S} r_i$ different values which can then be modeled
with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
necessarily define a distribution for $D$, but it is easy to verify
that it coincides with the $P_{NML}(D;G)$ for all the networks that
are composed of fully connected components.  The number of such
networks equals the number of nonempty partitions of a set of $n$
elements, i.e., the $n^\text{th}$ Bell number.

@@

The qNML yields equal scores for network structures that encode same
sets of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those who a) are same when directed
arcs are substituted by undirected ones and b) who have same
V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
$B$ are parents of $C$, but there is no arc between $A$ and $B$ (in
neither direction).  Later, Chickering~\cite{Chick95} showed that all
the equivalent network structures, and only them, can be reached from
each other by reversing, one by one, so called covered arcs, i.e. the
arcs from node $A$ to $B$, for which $B$'s parents other than $A$ are
exactly the $A$'s parents ($G_B=\{A\}\cup G_A$).

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
supplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notations and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  due to equations
  $\{A\}\cup G_A  = G_B$,
  $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
  and $G_A        = G'_B$.
\end{proof}

@@

The BIC can be written as
\begin{equation}
\textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
\end{equation}
where $D_i = D[\cdot, i]$ stands for the data on variable $i$,
$\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of
the conditional distribution of variable $i$ given its parents in
$G$. Since the both BIC and qNML scores are decomposable, we can focus
on studying the local scores. Using the definition of qNML, we have
\begin{align}\label{qnmlDef2}
s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}
  )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\ & -(reg(N,q_i
r_i) - reg(N,q_i)).
\end{align}

@@

\subsection{qNML equals NML for exponentially many models}
The fNML criterion equals NML criterion just for the Bayesian network
with no arcs.  We will next show that the qNML equals NML when the
connected components of the network are tournaments (complete directed
acyclic subgraphs of G).

We first note that the NML-criterion for the Bayesian networks
decomposes by the connected components. Namely, if $G$ consists of $C$
connected components $(G^1,\ldots,G^C)$ with variable sets $(V^1,\ldots,V^C)$,
the maximum likelihood for the data $D$ decomposes, and we can write
\begin{eqnarray}
  \lefteqn{P_{NML}(D;G)} \nonumber \\
  && = \frac{P(D;\hat\theta(D),G)}
            {\sum_{D_{V_1}}\ldots \sum_{D_{V_C}}\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} \nonumber \\
            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
            {\prod_{c=1}^C\sum_D^cP(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
\end{eqnarray}

Clearly, the qNML score also decomposes by the connected components,
so it remains to show that if the (sub)network $G$ is a tournament,
then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
score equivalence of both the NML criterion and the qNML criterion, we
may pick such a tournament $G$ that linear ordering defined by $G$
matches the ordering of the data columns, i.e., $i<j$ implies $G_i
\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
criterion we see that for the tournament $G$, the sum telescopes
leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.
This follows, since for any data vector $x$ in data $D$, the
$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$,
where we have denoted by $P^1$ the model that takes $n$-dimensional
vectors to be values of the single (collapsed) categorical variable.
Denoting prefixes of data vector $x$ by $x^{:i}$, and number of times
such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
($n_D(x^0)=N)$, we have
\begin{eqnarray}
  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\theta(D),G) \nonumber \\
  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
  &=& \prod_{j=1}^{N} \frac{n_D(d_j^{:n})}{N} \nonumber \\
  &=& P^1(D;\hat\theta^1(D),G).
\end{eqnarray}
Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
maximum likelihood probabilities, the equality above implies the equality
of these distributions.

Equality established, we would like to still state the number of
different networks connected components of which are tournaments.  We
start by generating all the $p(n)$ integer partitions i.e. ways to
partition $n$ labelled into parts with different size-profiles.  For
example, for $n=4$, we have $p(4)=5$ partition profiles $[[4], [3, 1],
  [2, 2], [2, 1, 1], [1, 1, 1, 1]]$.  Each of these partition size
profiles corresponds to many different networks, apart from the last
one that corresponds just to the empty network.  We count number of
networks for one such partition size profile (and then later sum these
counts up).  For any such partition profile $(p_1,\ldots,p_k)$ we can
(think we can) count the ways we can pick the nodes to different parts
and then order each part. This leads to the product $n\choose
p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
p_3$$p3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
the order of different parts of the same size does not matter, so for
all groups of parts having the same size, we have to divide the
product above by the factorial of the size of such group. Notice also,
that the product above telescopes, leaving us a formula for OEIS
sequence A000262 as described by Thomas Wieder:

With $p(n) =$ the number of integer partitions of $n$, 
     $d(i)$ = the number of different parts of the $i$-th partition of $n$, $m(i,j) =$ multiplicity of the $j$-th part of the $i$-th partition of $n$,  one has:

$$
a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
$$

HERE SOME TEXT ABOUT GROWTH RATE
